Supplementary Materials

### 2023-06-13 manuscript/manuscript.tex
ms r2 sec

Along the chromosome, our approach models the observed megabase-scale diversity
in window $i$ as the predicted level under our negative selection model,
$B_i'$, plus some residual error $\varepsilon_i$, $\pi_i = B_i' +
\varepsilon_i$. The out-sample $R_\text{LOO}^2$ estimates the ratio of variance
explained by the negative selection model ($\var_i(\pi_i)$) to the total
variance in diversity along the genome ($\var_i(\pi_i) = \var_i(B_i) +
\var_i(\varepsilon_i) + 2\cov_i(B_i', \varepsilon_i)$, \cite{Murphy2022-sj}).
Since our $R_\text{LOO}^2$ estimates suggest our models fit left-out
chromosomes well even though our method assumes constant demography and
homogeneous mutation rates along the genome, we wondered what $R^2$ we could
expect under a simple theoretic model where $\cov_i(B_i', \varepsilon_i) = 0$
and $\var_i(\varepsilon_i)$ was determined by the variance in coalescence times
due to drift alone with an effective population size rescaled by $B_i'$. We
calculate this ``drift $R^2$" by calculating the theoretic coalescence time
variance in a megabase window with recombination and an effective population
size of $B_i' N_e$ where we use our best-fitting model's as a plug-in estimator
of $B_i'$. We note that this is a rough approximation, since this calculation
assumes constant demography and that the variance in coalescence times under
negative selection is likely lower than expected under drift alone. Still, we
find that our out-sample $R_\text{LOO}^2$ for Yoruba individuals
($R_\text{LOO}^2 \approx 68$\%) is close to the theoretic drift
$R_\text{drift}^2 \approx 67$\%. For bottlenecked out-of-Africa populations,
$R_\text{drift}^2 \approx 64$\% compared to the observed $R_\text{LOO}^2
\approx 62$\% for CEU and $R_\text{LOO}^2 \approx 59$\% for CHB. We note that
because bottlenecks increase the variance in coalescence times beyond the level
implied by the effective population size, the disagreement between observed
$R_\text{LOO}^2$ and the theoretic $R_\text{drift}^2$ is likely to shrink.
Overall, this suggests that negative selection models explain the vast majority
coalescence time variation at the megabase-scale that is capable of being
explained (i.e. that is not coalescence noise).

>>>>

Since our $R_\text{LOO}^2$ estimates suggest our models fit left-out
chromosomes well even though our method assumes constant demography and
homogeneous mutation rates along the genome, we wondered what $R^2$ we could
expect under a simple theoretic model where $\cov_i(B_i', \varepsilon_i) = 0$
and $\var_i(\varepsilon_i)$ was determined by the variance in coalescence times
due to drift alone with an effective population size rescaled by $B_i'$. We
calculate this ``drift $R^2$" by calculating the theoretic coalescence time
variance in a megabase window with recombination and an effective population
size of $B_i' N_e$ where we use our best-fitting model's as a plug-in estimator
of $B_i'$. We note that this is a rough approximation, since this calculation
assumes constant demography and that the variance in coalescence times under
negative selection is likely lower than expected under drift alone.

Still, we find that our out-sample $R_\text{LOO}^2$ for Yoruba individuals
($R_\text{LOO}^2 \approx 68$\%) is close to the theoretic drift
$R_\text{drift}^2 \approx 67$\%. For bottlenecked out-of-Africa populations,
$R_\text{drift}^2 \approx 64$\% compared to the observed $R_\text{LOO}^2
\approx 62$\% for CEU and $R_\text{LOO}^2 \approx 59$\% for CHB. We note that
because bottlenecks increase the variance in coalescence times beyond the level
implied by the effective population size, the disagreement between observed
$R_\text{LOO}^2$ and the theoretic $R_\text{drift}^2$ is likely to shrink.
Overall, this suggests that negative selection models explain the vast majority
coalescence time variation at the megabase-scale that is capable of being
explained (i.e. that is not coalescence noise).

### 2023-06-13 manuscript/supplementary.tex
new ms stuff
ms r2 sec

\caption{$R^2$ and mutation rate estimates for all models}

@@

\section*{Approximate $R^2$ Limit}

We derive a rough approximation of the theoretic upper bound of $R^2$ by
noting that,

\begin{align}
    R^2 &= \frac{SS_\text{reg}}{SS_\text{reg} + SS_\text{res}} \\ 
        &= \frac{\sum_i (\pi_i - \widehat{\pi}_i)^2}{\sum_i (\pi_i - \widehat{\pi}_i)^2 + SS_\text{res}}
\end{align}

where $\pi_i$ is the observed diversity in window $i$, and $\widehat{\pi}_i$ is
the predicted diversity under a particular model.

Then, $SS_\text{res}$ is the residual noise due to the randomness in mutation
and average coalescence times in the window. If we make the assumptions that
mutation rates are heterogeneous across the genome and the coalescence rate in
a window is simply a background rate rescaled by the reduction factor $B_i$,
then,

\begin{align}
    SS_\text{res} &= \frac{1}{n} \sum_{i=1}^n \var(w\pi_i) \\ 
                  &= \frac{1}{n} \sum_{i=1}^n \var(w\pi_0 B_i) 
\end{align}

where $w$ is the width of the window such that the per-basepair $\pi$ is
appropriately scaled. We use the predicted $\hat{B}_i$ as the plugin estimator
for $B_i$. Then, we note that the variance in pairwise diversity is equivalent
to the variance in the number of segregating sites for a sample of two,
$\var(\pi) = \var(S_2)$. From \textcite[eq. 7.20]{Wakeley2009-ua}

\begin{align}
    \var(S_2) = \theta + \theta^2 \frac{2}{\rho^2} \int_0^\rho (\rho - x) f_2(x) dx
\end{align}
where,
\begin{align}
    f_2(\rho) = \frac{\rho + 18}{\rho^2 + 13 \rho + 18}.
\end{align}

Since this is a rough approximation, we set $\theta = w \pi_0 B_i$ and fix
$\rho = w \gamma \pi_0 B_i$ where $\gamma$ is the ratio of recombination to
mutation rates, $\nicefrac{r}{\mu}$. In humans, $\gamma \approx 1$, but explore
different ratios to assess sensitivity of this calculation (Supplementary
Figure \ref{suppfig:r2_upper}).

>>>>

\caption{$R^2$ and mutation rate estimates for all models. XXX note about repeats}

@@

\section*{Derivation of Approximate Drift $R^2$}

Our approach models the observed megabase-scale diversity $y_i$ in window $i$
as the predicted level under our negative selection model, $\pi_0 B_i$, plus
some random residual error $\varepsilon_i$,

\begin{align}
    y_i = \pi_0 B_i + \varepsilon_i
\end{align}

where each $B_i = f(X_i | \Psi)$ for some data $X_i$ (i.e. the recombinaton map
and annotation model) and parameters $\Psi$. To compare our models and assess
model goodness-of-fit, we use the out-sample $R_\text{LOO}^2$ which is
calculated by fitting the genome-wide model leaving one chromosome out, and
calculating the residual variance between predictions and observed values for
this out-sample chromosome.

In general, $R^2$ is calculated for a model fit across windows as,

\begin{align}
    R^2 = 1 - \frac{V_\text{res}}{V_\text{tot}}
\end{align}

where $V_\text{res}$ is the mean squared error across windows and
$V_\text{tot}$ is the total variance in the predicator. Suppose we have
estimates $\widehat{\pi}_0 \widehat{B}_i$ for $\pi_0 B_i$ from our model. Then,

\begin{align}
    V_\text{res} &= \frac{1}{n} \sum_{i=1}^n (y_i - \widehat{\pi}_0 \widehat{B}_i)^2  \\
                 &= \underbrace{\frac{1}{n} \sum_{i=1}^n \varepsilon_i^2}_\text{irreducible error} + \underbrace{\left(\frac{1}{n} \sum_{i=1}^n (\pi_0 B_i - \widehat{\pi}_0 \widehat{B}_i)\right)^2}_\text{bias squared}.
\end{align}

The total variance in the predictor is,

\begin{align}
    V_\text{tot} = V_\text{res} + V_\text{model} + 2\widehat{\pi}_0\cov_i(\varepsilon_i, \widehat{B}_i)
\end{align}
where 
\begin{align}
    V_\text{model} = \frac{\widehat{\pi}_0^2}{n} \sum_{i=1}^n \left(\widehat{B}_i - \frac{1}{n}\sum_{i=1}^n\widehat{B}_i\right)^2.
\end{align}

Generally, fitting procedures act to minimize the squared deviation between
predictors and observed values, which drives $\cov_i(\varepsilon_i,
\widehat{B}_i)$ to zero XXX. \textcite{Murphy2022-sj} suggest  XXX

Thus, our $R^2$ depends on (1) the bias of our predictors, (2) the covariance
between residuals and predictors, and (3) the irreducible error due to variance
in diversity within windows. However, it is of interest to approximate the
irreducible error under theoretic models that assume all irreducible error is
due entirely to the variance of coalescence times in a window, assuming that
all of the effects of negative selection on the variance can be thought of as a
local reduction in the effective population size to $B_i N_e$ and demographic
effects amount to a simple rescaling of $N_e$. This also assumes mutation rates
are constant across the genome, and do not contribute to the irreducible
variance. We call this theoretic goodness-of-fit under drift
$R_\text{drift}^2$, and it provides a rough approximation of the irreducible
error in our model due to ``coalescence noise" around the modeled reductions
due to negative selection.

Then,

\begin{align}
    \E(\varepsilon_i^2) = \var(\pi_i)
\end{align}

since the mean residual is zero. The variance in pairwise diversity
$\var(\pi_i)$ is equivalent to the variance in the number of segregating sites
for a sample of two, $\var(\pi) = \var(S_2)$. The variance in coalescence times
in a window of width $w$ basepairs with population-scaled recombination rate
$\rho = 4 B_i N_e r w$ is,

\begin{align}
    \var(S_2) = \theta + \theta^2 \frac{2}{\rho^2} \int_0^\rho (\rho - x) f_2(x) dx
\end{align}
(\cite[eq. 7.20]{Wakeley2009-ua}) where,
\begin{align}
    f_2(\rho) = \frac{\rho + 18}{\rho^2 + 13 \rho + 18}.
\end{align}

Since this is a rough approximation,  we set $\theta = w \pi_0 B_i$ and fix
$\rho = w \gamma \pi_0 B_i$ where $\gamma$ is the ratio of recombination to
mutation rates, $\nicefrac{r}{\mu}$. In humans, $\gamma \approx 1$, but we
explore different ratios to assess sensitivity of this calculation
(Supplementary Figure \ref{suppfig:r2_upper}).
